Simulating High-Dimensional Multivariate Data

### 2021-06-25 000-introduction.Rmd
Submitting to JSS, final pre-submission edits

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.

As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19]. For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with existing software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the HD random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependence perfectly. However, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, implement Pearson matching, but several layers of approximation are required.

>>>>

Massive high-dimensional (HD) data sets are now common in many areas of scientific inquiry. As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.

As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19]. For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied and implemented in publicly-available software [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations --- such as admitting only positive correlations [@Park1996]. While more recent approaches [@MB13; @Xia17; @BF17] have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the HD random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc.). For normal data, the Pearson product-moment correlation describes the dependence perfectly. However, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties enabling our proposed methodology. We do, however, implement Pearson matching, but several layers of approximation are required.

### 2021-06-25 010-background.Rmd
Submitting to JSS, final pre-submission edits

* SP1: Procedure must scale to high dimensions with practical compute times.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

@@

To place `Bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.  In `R`, the `MASS` package's `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This procedure cannot simulate from arbitrary margins and dependency measures and it fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks. Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released, which provides HD multivariate normal generation. However, lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD `R` packages exist, such as `GenOrd`. For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vector generation via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table \@ref(tab:compare-table) summarizes the above discussion.

>>>>

* SP1: Procedure must scale to high dimensions with practical computation times.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

@@

To place `Bigsimr`'s implementation and algorithms in context of existing approaches, we qualitatively compare `Bbigsimr` to other multivariate simulation packages.  In `R`, the `MASS` package's `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This procedure cannot simulate from arbitrary margins and dependency measures and it fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks. Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released, which provides HD multivariate normal generation. However, lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD `R` packages exist, such as `GenOrd`. For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vector generation via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence structure. Table \@ref(tab:compare-table) summarizes the above discussion.

### 2021-06-25 020-simulation-algorithm.Rmd
Submitting to JSS, final pre-submission edits

This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$. Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$. To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$. The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a HD NORTA-inspired algorithm.

@@

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated. Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75]. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix becomes non-positive definiteness. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@@

There is hope, however, as @GH02 also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.

@@

There is an increasing probability of encountering an inadmissible correlation matrix as dimension increases. In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix. In Section \@ref(package) we provide a basic description of how to quantify and control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

>>>>

This section describes our methods for simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$. Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$. To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$. The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a HD NORTA-inspired algorithm.

@@

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated. Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75]. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined, the resultant matrix becomes non-positive definite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@@

There is hope, however, as @GH02 also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides evidence that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.

@@

There is an increasing probability of encountering an inadmissible correlation matrix as dimension increases. In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix. In Section \@ref(package), we provide a basic method of how to quantify and control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

### 2021-06-25 040-monte-carlo-evaluation.Rmd
Submitting to JSS, final pre-submission edits

In this section, we conduct MC studies to investigate method performance. Marginal parameter matching is straightforward in our scheme as it is a sequence of univariate inverse probability transforms. This process is both fast and exact. On the other hand, accurate and computationally efficent dependency matching presents a challenge, especially in the HD setting. To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.

@@

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example. Thus, we proceed in increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data. We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations. For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for each correlation type, to evaluate the simulation's ability to simulate all possible correlations. The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).

Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations. We increase the simulate replicates $B$ to visually ensure that our results converge to the target correlations and gauge efficiency. We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types. Then, we select a non-symmetric continuous distribution: a standard (rate =1), two-component multivariate gamma (MVG). Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [RNA-seq data application](examples) for estimation details). Specifically, we assume a MVNB with $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.

>>>>

In this section, we conduct MC studies to investigate method performance. Marginal parameter matching is straightforward in our scheme as it is a sequence of univariate inverse probability transforms. This process is both fast and exact. On the other hand, accurate and computationally efficient dependency matching presents a challenge, especially in the HD setting. To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.

@@

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example. Thus, we proceed by increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data. We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations. For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for each correlation type, to evaluate the simulation's ability to generate all possible correlations. The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).

Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations. We increase the simulation replicates $B$ to visually ensure that our results converge to the target correlations and gauge efficiency. We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types. Then, we select a non-symmetric continuous distribution: a standard (rate =1), two-component multivariate gamma (MVG). Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [RNA-seq data application](examples) for estimation details). Specifically, we assume a MVNB with $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.

### 2021-06-25 070-supplementary-materials.Rmd
Submitting to JSS, final pre-submission edits

We provide an open-source implementation of our methodology as the `Bigsimr.jl` Julia package, hosted on github at https://github.com/adknudson/Bigsimr.jl. Additionally we provide R and Python interfaces to `Bigsimr`, respectively on https://github.com/SchisslerGroup/r-bigsimr and https://github.com/SchisslerGroup/python-bigsimr. A stable release of the R package `bigsimr` is also available on CRAN. `Bigsimr.jl` is an ongoing project, and feature requests or issues can be submitted to https://github.com/adknudson/Bigsimr.jl/issues.

>>>>

We provide an open-source implementation of our methodology as the `Bigsimr.jl` Julia package, hosted on github at https://github.com/SchisslerGroup/Bigsimr.jl. Additionally we provide R and Python interfaces to `Bigsimr`, respectively on https://github.com/SchisslerGroup/r-bigsimr and https://github.com/SchisslerGroup/python-bigsimr. A stable release of the R package `bigsimr` is available on CRAN. `Bigsimr.jl` is an ongoing project, and feature requests or issues can be submitted to https://github.com/adknudson/Bigsimr.jl/issues.

### 2021-06-25 index.Rmd
Submitting to JSS, final pre-submission edits

---
documentclass: jss
author:
  - name: Alfred G. Schissler
    affiliation: University of Nevada
    address: |
      | 1664 N Virginia St.
      | Reno, NV 89557
    email: \email{aschissler@unr.edu}
  - name: Edward J. Bedrick
    affiliation: University of Arizona
  - name: Alexander D. Knudson
    affiliation: University of Nevada \AND
  - name: Tomasz J. Kozubowski
    affiliation: University of Nevada
  - name: Tin Nguyen
    affiliation: University of Nevada
  - name: Anna K. Panorska
    affiliation: University of Nevada \AND
  - name: Juli Petereit
    affiliation: University of Nevada
  - name: Walter W. Piegorsch
    affiliation: University of Arizona
  - name: Duc Tran
    affiliation: University of Nevada
title:
  formatted: "Simulating High-Dimensional Multivariate Data using the \\pkg{Bigsimr} Package"
  plain:     "Simulating High-Dimensional Multivariate Data using the Bigsimr Package"
  short:     "\\pkg{Bigsimr}: Simulate High-Dimensional Multivariate Data"
abstract: >
  It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. To meet this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces. These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix. \pkg{Bigsimr} contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix. Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$. We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
keywords:
  formatted: [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
  plain:     [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
bibliography: [bigsimr.bib, packages.bib, alex.bib]
biblio-style: jss
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: preamble.tex
---

>>>>

---
documentclass: jss
author:
  - name: Alfred G. Schissler
    affiliation: University of Nevada
    address: |
      | 1664 N Virginia St.
      | Reno, NV 89557
    email: \email{aschissler@unr.edu}
  - name: Edward J. Bedrick
    affiliation: University of Arizona
  - name: Alexander D. Knudson
    affiliation: University of Nevada \AND
  - name: Tomasz J. Kozubowski
    affiliation: University of Nevada
  - name: Tin Nguyen
    affiliation: University of Nevada
  - name: Anna K. Panorska
    affiliation: University of Nevada \AND
  - name: Juli Petereit
    affiliation: University of Nevada
  - name: Walter W. Piegorsch
    affiliation: University of Arizona
  - name: Duc Tran
    affiliation: University of Nevada
title:
  formatted: "Simulating High-Dimensional Multivariate Data using the \\pkg{Bigsimr} Package"
  plain:     "Simulating High-Dimensional Multivariate Data using the Bigsimr Package"
  short:     "\\pkg{Bigsimr}: Simulate High-Dimensional Multivariate Data"
abstract: >
  It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained in high-throughput biomedical experiments. Due to the computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. To close this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces. These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix. \pkg{Bigsimr} contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix. Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$. We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
keywords:
  formatted: [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
  plain:     [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
bibliography: [bigsimr.bib, packages.bib, alex.bib]
biblio-style: jss
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: preamble.tex
---
